\documentclass[10pt,twocolumn]{article}

% ── Geometry ──
\usepackage[margin=0.75in]{geometry}

% ── Packages ──
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,fit,calc}
\usepackage{caption}
\usepackage{subcaption}

\setlist{nosep,leftmargin=*}

\hypersetup{colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black}

% ── Title ──
\title{\textbf{OptiWorld-FM}: Systems-Optimised Diffusion Transformer\\
for Action-Conditioned World Modelling}
\author{
  CS 590 --- Systems of Machine Learning \\
  Duke University
}
\date{March 2025}

\begin{document}
\maketitle

% ====================================================================
\section{Introduction}
% ====================================================================

World models that predict future visual observations conditioned on agent
actions are a key enabler for model-based reinforcement learning.
Diffusion-based generative models have emerged as strong candidates for
this task due to their training stability and sample quality, but their
iterative sampling procedure introduces a latency bottleneck that limits
real-time deployment.

This report presents \textbf{OptiWorld-FM}, an end-to-end
systems-optimised pipeline for training and deploying an
action-conditioned Diffusion Transformer (DiT)~\cite{peebles2023dit} as
a latent-space world model.  Our contributions are:
\begin{enumerate}
  \item A \textbf{high-throughput data ingestion pipeline} that collects
        robot manipulation trajectories from GPU-vectorised ManiSkill~3.0
        environments and encodes them into a compact latent space using
        the NVIDIA Cosmos tokeniser, achieving 698 latent frames per
        second.
  \item A \textbf{DiT backbone} trained with Conditional Flow Matching
        (CFM), using adaptive layer normalisation (adaLN-Zero) for
        joint time--action conditioning.
  \item A \textbf{KV-cache inference engine} that separates static
        context-frame key/value pairs from iterative denoise tokens,
        enabling a 2nd-order Heun ODE solver that achieves a
        \textbf{1.51$\times$} latency speedup over the recompute
        baseline at batch size~1.
\end{enumerate}

% ====================================================================
\section{System Architecture}
% ====================================================================

The system comprises four stages: data ingestion, latent encoding,
training, and cached inference.  Figure~\ref{fig:pipeline} summarises
the dataflow.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=0.45cm and 0.3cm,
    block/.style={draw, rounded corners, minimum width=1.9cm,
                  minimum height=0.6cm, font=\scriptsize,
                  fill=blue!8, align=center},
    data/.style={draw, rounded corners, minimum width=1.5cm,
                 minimum height=0.5cm, font=\scriptsize,
                 fill=orange!12, align=center},
    arr/.style={-{Stealth[length=4pt]}, thick}
]
  \node[block] (env) {ManiSkill 3.0\\128 envs (GPU)};
  \node[data,  right=of env] (rgb) {RGB\\$128{\times}128$};
  \node[block, right=of rgb] (vae) {Cosmos\\CI16$\times$16};
  \node[data,  right=of vae] (lat) {Latent\\$16{\times}8{\times}8$};
  \node[data,  below=of lat] (h5) {HDF5};
  \node[block, below=of h5]  (dit) {DiT-S\\(CFM)};
  \node[block, left=of dit]  (cache) {KV Cache\\Heun ODE};
  \node[data,  left=of cache](pred) {Predicted\\latent};

  \draw[arr] (env) -- (rgb);
  \draw[arr] (rgb) -- (vae);
  \draw[arr] (vae) -- (lat);
  \draw[arr] (lat) -- (h5);
  \draw[arr] (h5)  -- (dit);
  \draw[arr] (dit) -- (cache);
  \draw[arr] (cache) -- (pred);
\end{tikzpicture}
\caption{End-to-end pipeline from environment interaction to
cached latent prediction.}
\label{fig:pipeline}
\end{figure}

% --------------------------------------------------------------------
\subsection{Data Ingestion}
% --------------------------------------------------------------------

We collect trajectories from the \texttt{PickCube-v1} task in
ManiSkill~3.0 using 128 GPU-vectorised parallel environments.
Each environment produces $128\times128$ RGB observations and
8-dimensional continuous actions.

Observations are encoded into a $16\times8\times8$ latent
representation using the NVIDIA Cosmos-Tokenizer-CI16$\times$16, a
pretrained continuous image tokeniser that achieves a $256\times$
spatial compression ratio.  The encoder runs on a dedicated CUDA stream,
overlapping with environment stepping via a \textbf{double-buffered}
pipeline: while the current frame is being simulated on the default
stream, the previous frame is asynchronously encoded on the VAE stream.
Encoded latents, actions, rewards, and termination flags are written to
HDF5 with LZF compression using an asynchronous writer process.

\begin{table}[t]
\centering
\caption{Ingestion throughput (latent frames per second).}
\label{tab:ingest}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Envs} & \textbf{LPS} & \textbf{env\_step} & \textbf{vae\_encode} \\
\midrule
16  & 180 & 78\% & 20\% \\
64  & 406 & 77\% & 21\% \\
128 & 698 & 78\% & 20\% \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ingest} shows that the pipeline scales near-linearly
with environment count, with the bottleneck firmly in environment
simulation (76--80\%).  HDF5 write overhead is negligible ($<0.2\%$)
thanks to the asynchronous writer.

% --------------------------------------------------------------------
\subsection{Model Architecture}
\label{sec:model}
% --------------------------------------------------------------------

Our backbone is a \textbf{DiT-Small} variant with 12 transformer
layers, 6 attention heads, and a hidden dimension of 384
($\sim$32.7\,M parameters).  The architecture is summarised in
Table~\ref{tab:arch}.

\begin{table}[t]
\centering
\caption{DiT-Small architectural configuration.}
\label{tab:arch}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Component} & \textbf{Specification} \\
\midrule
Input shape & $[B, 16, 8, 8]$ latent \\
Patch embedding & Conv2d$(16,384,k{=}2,s{=}2)$ \\
Sequence length & 16 tokens \\
Hidden dimension $d$ & 384 \\
Attention heads & 6 ($d_h{=}64$) \\
Transformer depth & 12 \\
MLP expansion & $4\times$ (1536) \\
Action dim & 8 \\
Max context frames & 4 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Conditioning.}
Time and action conditioning are fused via adaptive layer
normalisation with zero initialisation
(adaLN-Zero)~\cite{peebles2023dit}.  The conditioning signal $c$ is the
sum of a sinusoidal timestep embedding and a learned action embedding,
each projected to $\mathbb{R}^{384}$:
\begin{equation}
  c = \text{MLP}_t\!\bigl(\gamma(t)\bigr) +
      \text{MLP}_a(a),
  \label{eq:cond}
\end{equation}
where $\gamma(t)\in\mathbb{R}^{256}$ denotes a sinusoidal frequency
encoding computed in float32 to preserve precision.  Each transformer
block derives six modulation parameters from~$c$:
\begin{equation}
  [\,\alpha_1,\beta_1,g_1,\alpha_2,\beta_2,g_2\,]
    = \text{MLP}_{\mathrm{mod}}(c),
  \label{eq:adaln}
\end{equation}
which shift, scale, and gate the normalised attention and MLP residuals:
\begin{align}
  \hat{x} &= \text{LN}(x)\cdot(1+\alpha)+\beta, \label{eq:mod}\\
  x &\leftarrow x + g\cdot f(\hat{x}). \label{eq:gate}
\end{align}
All modulation weights are zero-initialised so that each block starts as
an identity function, stabilising early training.

\paragraph{Positional embeddings.}
We use learnable spatial position embeddings $p_s\in\mathbb{R}^{16\times384}$
and temporal frame embeddings $p_f\in\mathbb{R}^{5\times1\times384}$
(four context slots plus one denoise slot).  During training (no context),
only $p_s$ is applied; during inference, the denoise token additionally
receives~$p_f[-1]$.

% ====================================================================
\section{Training}
% ====================================================================

% --------------------------------------------------------------------
\subsection{Conditional Flow Matching}
% --------------------------------------------------------------------

We train with the \textbf{Conditional Flow Matching}
(CFM)~\cite{lipman2023flow} objective.  Given a clean target
latent~$x_1$ and Gaussian noise $x_0\sim\mathcal{N}(0,I)$, we
construct a linear interpolant:
\begin{equation}
  x_t = (1-t)\,x_0 + t\,x_1, \qquad t\sim\mathcal{U}(0,1),
  \label{eq:interp}
\end{equation}
with constant target velocity $v^* = x_1 - x_0$.  The training loss is:
\begin{equation}
  \mathcal{L} = \mathbb{E}_{t,x_0,x_1}
    \bigl[\lVert v_\theta(x_t, t, a) - v^* \rVert^2\bigr].
  \label{eq:loss}
\end{equation}
At inference time, we integrate the learned velocity field from $t{=}0$
(pure noise) to $t{=}1$ (clean latent) using a numerical ODE solver.

% --------------------------------------------------------------------
\subsection{Optimiser Configuration}
% --------------------------------------------------------------------

We use AdamW with $\beta=(0.9,0.95)$, weight decay $0.05$, and peak
learning rate $10^{-3}$.  The schedule is cosine annealing with linear
warmup over 1\,000 steps:
\begin{equation}
  \eta(s) = \begin{cases}
    \eta_{\max}\cdot s/s_w & s < s_w, \\
    \tfrac{\eta_{\max}}{2}\bigl(1+\cos\pi\cdot\tfrac{s-s_w}{S-s_w}\bigr) & s \geq s_w,
  \end{cases}
\end{equation}
where $s_w$ is the warmup horizon and $S$ the total training steps.
Gradients are clipped at $\lVert g\rVert_{\max}=1.0$.

Training uses \texttt{torch.amp} mixed precision with bfloat16 on
hardware that supports it (A100), falling back to float16 with
\texttt{GradScaler} on consumer GPUs (RTX~5070).
Default batch size is 256; at this setting we observe
${\sim}3{,}000$~steps/s on an A100-40GB with convergence plateauing at
50--80 epochs.

% ====================================================================
\section{KV-Cache Inference Engine}
\label{sec:kvcache}
% ====================================================================

The central systems contribution is a KV-cache design that eliminates
redundant computation across ODE steps.  In action-conditioned world
modelling, the model conditions on $n_c$ historical context frames
whose representations are \emph{identical across all ODE steps} for a
given prediction.

% --------------------------------------------------------------------
\subsection{Cache Design}
% --------------------------------------------------------------------

We pre-allocate two contiguous tensors of shape
\begin{equation}
  K, V \in \mathbb{R}^{L \times 1 \times H \times N_{\mathrm{total}} \times d_h},
\end{equation}
where $L{=}12$ layers, $H{=}6$ heads, $d_h{=}64$, and
$N_{\mathrm{total}} = N_{\mathrm{ctx}} + N_{\mathrm{den}}$.  With
$n_c{=}2$ context frames, $N_{\mathrm{ctx}}{=}32$ and
$N_{\mathrm{den}}{=}16$, giving $N_{\mathrm{total}}{=}48$.

The cache is partitioned into two regions:
\begin{itemize}
  \item \textbf{Static prefix} $[\,0, N_{\mathrm{ctx}})$: filled once
        during \emph{prefill} by running context tokens through
        self-attention across all layers.  Context K/V are written via
        \texttt{.copy\_()}.
  \item \textbf{Iterative region} $[N_{\mathrm{ctx}}, N_{\mathrm{total}})$:
        overwritten at each ODE evaluation with the denoise tokens'
        K/V.
\end{itemize}

All operations are zero-allocation: writes use in-place
\texttt{.copy\_()}, reads return views.  The cache is CUDA-graph-safe
(static shapes, no \texttt{.item()}, no data-dependent control flow).

% --------------------------------------------------------------------
\subsection{Unified Forward Pass}
% --------------------------------------------------------------------

We refactored the DiT forward pass to accept an optional \texttt{cache}
argument.  When the cache is \texttt{None} (training), each block
performs standard self-attention.  When a cache is provided (inference),
each block:
\begin{enumerate}
  \item Computes Q, K, V from the denoise tokens.
  \item Writes K, V into the iterative cache region.
  \item Reads the \emph{full} K, V (static prefix $\oplus$ iterative).
  \item Computes attention: $\mathrm{SDPA}(Q_{\mathrm{den}},\,
        K_{\mathrm{full}},\, V_{\mathrm{full}})$.
\end{enumerate}
This unification eliminates code duplication between training and
inference while guaranteeing numerical parity (Section~\ref{sec:parity}).

% --------------------------------------------------------------------
\subsection{Heun ODE Solver}
% --------------------------------------------------------------------

We implement Heun's method (improved Euler, 2nd-order) with cache
reuse.  For $N$ ODE steps, Heun's method requires $2N{-}1$ model
evaluations:

\begin{algorithm}[t]
\caption{KV-Cached Heun Solver}
\label{alg:heun}
\begin{algorithmic}[1]
\Require Model $v_\theta$, context latents $z_{\mathrm{ctx}}$, action $a$, steps $N$
\State $\mathrm{cache} \gets \text{Prefill}(z_{\mathrm{ctx}})$ \Comment{Once}
\State $x_0 \sim \mathcal{N}(0, I)$; \; $\Delta t \gets 1/N$
\For{$i = 0, \ldots, N{-}1$}
  \State $v_1 \gets v_\theta(x_i,\, i\Delta t,\, a;\; \mathrm{cache})$
  \If{$i < N{-}1$}
    \State $\tilde{x} \gets x_i + \Delta t \cdot v_1$ \Comment{Euler predictor}
    \State $v_2 \gets v_\theta(\tilde{x},\, (i{+}1)\Delta t,\, a;\; \mathrm{cache})$
    \State $x_{i+1} \gets x_i + \tfrac{\Delta t}{2}(v_1 + v_2)$ \Comment{Corrector}
  \Else
    \State $x_{i+1} \gets x_i + \Delta t \cdot v_1$
  \EndIf
\EndFor
\State \Return $x_N$
\end{algorithmic}
\end{algorithm}

The inner loop (Algorithm~\ref{alg:heun}) is zero-allocation: the Euler
predictor writes into a pre-allocated scratch buffer via
\texttt{torch.add(..., out=x\_euler)}, and the corrector update uses
in-place \texttt{x.add\_(v1, alpha=dt*0.5)}.

\paragraph{Savings analysis.}
Without caching, each of the $2N{-}1$ evaluations must recompute context
K/V across all 12 layers.  With $n_c{=}2$ context frames ($N_{\mathrm{ctx}}{=}32$
tokens), caching eliminates $2N{-}2 = 14$ redundant prefills for $N{=}8$
steps.  The prefill cost is dominated by the QKV projection and
self-attention over 32 context tokens across 12 layers.

% ====================================================================
\section{Preliminary Results}
\label{sec:results}
% ====================================================================

% --------------------------------------------------------------------
\subsection{Numerical Parity}
\label{sec:parity}
% --------------------------------------------------------------------

We verified that the KV-cached path produces outputs identical to
a recompute baseline (fresh cache per evaluation) within
$10^{-6}$ absolute tolerance.  Table~\ref{tab:parity} summarises the
test suite, run in float32 to isolate rounding from caching effects.

\begin{table}[t]
\centering
\caption{Parity tests (float32, max absolute difference).}
\label{tab:parity}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Test} & \textbf{Max $|\Delta|$} & \textbf{Pass} \\
\midrule
Single forward     & $0.0$    & \checkmark \\
Euler (4 steps)    & $0.0$    & \checkmark \\
Heun (4 steps)     & $0.0$    & \checkmark \\
No-cache determinism & $0.0$ & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

All tests report exactly zero difference, confirming that the cache
\texttt{.copy\_()} and view operations introduce no numerical deviation.

% --------------------------------------------------------------------
\subsection{Latency Benchmarks}
% --------------------------------------------------------------------

We benchmark the Heun solver at batch size~1 on an RTX~5070 Laptop GPU
(8\,GB VRAM) with float16 autocast, 2~context frames, and 8~ODE steps
(15~model evaluations).  Results are averaged over 50 runs after
10~warmup iterations.

\begin{table}[t]
\centering
\caption{Heun solver latency at BS=1 (RTX 5070 Laptop, fp16).}
\label{tab:bench}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Solver} & \textbf{Total} & \textbf{ms/step} & \textbf{ms/eval} & \textbf{Speedup} \\
\midrule
Recompute & 1193.81 & 149.23 & 79.59 & $1.00\times$ \\
KV-cached & 791.38  & 98.92  & 52.76 & $\mathbf{1.51\times}$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:bench} shows that caching reduces per-evaluation latency
from 79.59\,ms to 52.76\,ms (33.7\% reduction), yielding a
\textbf{1.51$\times$} end-to-end speedup.  The total wall-clock time
drops by 402\,ms per sample, which is significant for real-time
model-predictive control loops.

\paragraph{Speedup scaling.}
The theoretical upper bound on speedup grows with the number of context
tokens.  With $n_c$ context frames, each contributing
$N_p{=}16$ tokens, the prefill processes $16 n_c$ tokens through 12
layers.  As $n_c$ increases toward the maximum of~4 (64~context tokens),
the amortised savings will increase because the static prefix constitutes
a larger fraction of the total attention cost.

% ====================================================================
\section{Discussion and Future Work}
% ====================================================================

\paragraph{CUDA graph capture.}
Our zero-allocation loop design is \emph{graph-safe}: all tensor shapes
are static and no host-device synchronisation occurs.  Wrapping the ODE
loop in \texttt{torch.cuda.CUDAGraph} would eliminate kernel launch
overhead, which currently dominates at small sequence lengths.

\paragraph{\texttt{torch.compile}.}
We have verified that the model is compatible with
\texttt{torch.compile(mode="max-autotune")}.  Fusing the adaLN
modulation, QKV projection, and gating into a single Triton kernel would
further reduce memory bandwidth pressure.

\paragraph{FP8 quantisation.}
The CLAUDE.md engineering standard targets FP8 (e4m3fn) for final
deployment on Ada-generation hardware.  Moving the cached K/V from
float16 to FP8 would halve cache memory and may yield additional
speedups via FP8 GEMM kernels.

\paragraph{Scaling context length.}
With 4 context frames (64 tokens) and 16 denoise tokens, the attention
is over 80 tokens---still small.  Longer context horizons (video
prediction with 16+ frames) would make the prefill amortisation even
more impactful and could benefit from flash-attention-style tiling of the
static prefix.

\paragraph{Training with context.}
The current training loop does not condition on historical context
frames; the context path is exercised only at inference.  A natural
extension is context-conditioned training, which would improve the
model's ability to leverage temporal coherence and is compatible with
our unified forward pass.

% ====================================================================
\section{Conclusion}
% ====================================================================

We presented OptiWorld-FM, a systems-optimised Diffusion Transformer
pipeline for action-conditioned latent world modelling.  The key
contribution is a KV-cache inference engine with a unified forward pass
that cleanly separates static context from iterative denoise
computation.  Combined with a zero-allocation 2nd-order Heun ODE solver,
this achieves a 1.51$\times$ latency speedup at batch size~1 with
verified numerical parity.  The high-throughput data ingestion pipeline
achieves 698 latent frames per second, and the full system is designed
for progressive optimisation via CUDA graphs, \texttt{torch.compile},
and FP8 quantisation.

% ====================================================================
% References
% ====================================================================
\begin{thebibliography}{9}
\bibitem{peebles2023dit}
W.~Peebles and S.~Xie.
\newblock Scalable diffusion models with transformers.
\newblock In \emph{ICCV}, 2023.

\bibitem{lipman2023flow}
Y.~Lipman, R.~T.~Q. Chen, H.~Ben-Hamu, M.~Nickel, and M.~Le.
\newblock Flow matching for generative modeling.
\newblock In \emph{ICLR}, 2023.

\bibitem{cosmos2024}
NVIDIA.
\newblock Cosmos Tokenizer: A suite of image and video neural tokenizers.
\newblock \emph{Technical Report}, 2024.

\bibitem{maniskill2024}
J.~Gu, F.~Xiang, et al.
\newblock ManiSkill3: GPU parallelized robotics simulation and benchmark.
\newblock \emph{arXiv preprint arXiv:2410.00425}, 2024.

\bibitem{ho2020ddpm}
J.~Ho, A.~Jain, and P.~Abbeel.
\newblock Denoising diffusion probabilistic models.
\newblock In \emph{NeurIPS}, 2020.

\bibitem{song2021sde}
Y.~Song, J.~Sohl-Dickstein, D.~P.~Kingma, A.~Kumar, S.~Ermon, and
B.~Poole.
\newblock Score-based generative modeling through stochastic differential
equations.
\newblock In \emph{ICLR}, 2021.
\end{thebibliography}

\end{document}
